Learning pleiotropic signatures to construct a causal map of the human phenome

### 2017-07-29 docs/manuscript.rmd
updated manuscript

Gibran Hemani$^1$, Jack Bowden$^1$, Philip Haycock$^1$, Jie Zheng$^1$, Oliver Davis$^1$, Peter Flach, Tom Gaunt$^{1,*}$, George Davey Smith$^{1,*}$

1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this. Here we show that the mechanisms that underlie horizontal pleiotropy are numerous, and that instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over unselective use of any existing methods. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. All results are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).

@@

The use of summary data in two-sample MR is described in detail elsewhere [@Hemani2016]. A brief outline of the procedure is as follows. First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of MR (that the instrument associates with the exposure) is generally satisfied. We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait. These can be obtained manually from complete summary data or from curated lists of significant GWAS associations. Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because these SNPs will generally have not have reached genome-wide significance for other traits (and therefore be present in curated catalogues). At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome. This is known as the inverse-variance weighted (IVW) method.

@@

Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a). Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are likely to be exhibiting horizontal pleiotropy. Conversely, we primarily developed Steiger filtering to identify those instruments that are likely to be arising due to reverse cause.

The Steiger test is applied to each variant in turn and we exclude any $g_{A}$ for which $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$, indicating that it is unlikely to primarily associate with $B$ relative to $A$. Similarly, for SNPs that influence confounders of $A$ and $B$ or exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant. Hence we also exclude any $g_{A}$ for which the Steiger Z-test has $p > 0.05$.

To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$. If $x$ is binary then we estimate the variance of the underlying liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$. Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2p(1-p)$, where $\beta$ is the log odds ratio and $p$ is the allele frequency of the SNP in the population [@Lee2013c]. $p$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $p_{cc}, and number of cases $N_1$ and controls $N_0$. We exclude any SNPs that do not have the correct causal direction for the hypothesis, and do not have Steiger test $p < 0.05$.

@@

We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2). In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs. Our objective is to select the expert most likely to be correct for a specific MR analysis. A overview of the approach is shown in Figure 2.

The mixture of experts method is a machine learning approach [@Jordan1994] which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert. In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations. This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that dataset. The metrics are a mixture of regression diagnostics and are described in Supplementary table 3.

The gating function needs to be trained using data for which the true causal effect is known, and to this end we generated a large number of simulated datasets. We trained the gating function using random forest learning algorithms that seek to identify sectors of the parameter space that are most likely to return accurate estimates for a particular expert. Figure 2a illustrates how the gating function is trained using simulated data. New datasets can then be applied to the trained mixture of experts (MoE) model (Figure 2b). We call this implementation MR-MoE 1.0.

It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation. In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data. As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem. From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the datasets.

@@

The MoE is trained using datasets generated from simulations (Figure 2a). A *dataset* is the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure. This *dataset* can be fed into any of the 28 experts to obtain MR causal effects. The simulations used to generate these datasets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influence confounding variables.

We simulate two individual level datasets for which there are $N_x$ and $N_y$ samples, and $M$ SNPs, where each SNP has effect allele frequency of $p_m \sim U(0.05, 0.95)$. These datasets are used to obtain the SNP effects for the exposure trait $x$ and the outcome trait $y$, respectively, using the following sampling criteria:

@@

For each simulation we used linear regression to estimate the genetic effect of each SNP $M$ on $x$ in sample 1, and each SNP $M$ on $y$ in sample 2. We then perform MR analysis in both directions, mimicking GWAS by retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations). We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate datasets Hence, for each simulation two datasets are generated which are analysed to produce 28 MR estimates each. We performed 100,000 simulations using these parameters, resulting in 200,000 datasets.

@@

We aim to maximise statistical power for datasets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for datasets where $\beta_{x,y} = 0$. To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d})$ is generated where the training set of input metrics for dataset $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is

@@

where $p_{m,d}$ is the p-value for method $m$ on dataset $d$.

@@

For each training dataset we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated dataset. For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a dataset. To do this, for each method we train a random decision forest to predict that method's performance using the dataset's metrics. The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.

As a simple hypothetical example - if a dataset has a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:

@@

Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 datasets to predict which method would have the highest performance for each of the remaining datasets. Finally we compare the performance of the method selected by the MoE against all remaining methods. The default settings for the `randomForest` package in R [@Liaw2002] were used to train the models. MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR) [@Hemani2016].

The set of MR estimates obtained from this analysis are recorded in a Neo4j graph database. Because each association has up to 28 different estimates, for simplicity we distill this down to a single 'best estimate' for each association using the following rules:

As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a). We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated datasets comprising both null and non-null causal models. For each dataset, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.

Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular dataset based on its characteristics (Figure 2).

The ability to predict the performance for each of these methods is shown in Supplementary table 2. The prediction $R^2$ of whether a dataset's status was truly null ($\beta_{xy}=0$) or non-null ($\beta_{xy} \neq 0$) against the method's prediction of the dataset's status, ranged between 0.04 and 0.24. The method performance prediction was most effective for the Egger random effects model with Steiger filtering. The dataset characteristics with the most importance for each of the predictors differed substantially between each dataset, as well as the frequency for which each of the methods was selected in the testing datasets. The FDR of each method when chosen, compared to their averages across all datasets, reduced; and likewise the true positive rates of each method when chosen increased compared to their simulation-wide averages.

We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods. Figures 3a and 3b show that the MoE approach had the best general performance, obtaining an AUC of 0.84 in terms of classifying the simulations as being null or non-null. Notably, the next best methods were median and mode based estimators using Steiger filtering. Under the assumption of widespread pleiotropy it is clear that all methods suffer from high false discovery rates.

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception. Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific dataset. The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy. As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].

In addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic.

We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular dataset, namely, that the data is choosing the model. This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance. The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear. Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.

>>>>

Gibran Hemani$^1$, Jack Bowden$^1$, Philip Haycock$^1$, Jie Zheng$^1$, Oliver Davis$^1$, Peter Flach$^2$, Tom Gaunt$^{1,*}$, George Davey Smith$^{1,*}$

1. MRC Integrative Epidemiology Unit (IEU) at the University of Bristol, School of Social and Community Medicine, Bristol, UK
2. Intelligent Systems Laboratory, Department of Computer Science, University of Bristol

@@

A major application for genome-wide association studies (GWAS) has been the emerging field of causal inference using Mendelian randomization (MR), where the causal effect between a pair of traits can be estimated using only summary level data. MR depends on SNPs exhibiting vertical pleiotropy, where the SNP influences an outcome phenotype only through an exposure phenotype. Issues arise when this assumption is violated due to SNPs exhibiting horizontal pleiotropy, and many methods have been developed in an attempt to address this. Here we show that the mechanisms that underlie horizontal pleiotropy are numerous, and that instrument selection will be increasingly liable to selecting invalid instruments as GWAS sample sizes continue to grow. We have developed a mixture of experts machine learning approach (MR-MoE 1.0) that improves on both power and false discovery rates over unselective use of any existing methods. Using the approach, we systematically estimated the causal effects amongst 2407 phenotypes, generating a working draft of the causal map of the human phenome comprising the evaluation of causality between over a million pairs of traits. Almost 90% of causal estimates indicated some level of horizontal pleiotropy. The causal estimates are organised into a publicly available graph database ([http://eve.mrbase.org](http://eve.mrbase.org)).

@@

The use of summary data in two-sample MR is described in detail elsewhere [@Hemani2016]. A brief outline of the procedure is as follows. First, genetic instruments for the exposure trait need to be identified - those SNPs with $p < 10^{-8}$ are retained in order to ensure that the first assumption of MR (that the instrument associates with the exposure) is generally satisfied. We collect their effect sizes, standard errors and effect alleles for the association with the exposure trait. These can be obtained manually from complete summary data or from curated lists of significant GWAS associations. Next, the effects of those SNPs on the outcome need to be obtained, typically necessitating complete summary data because these SNPs will generally have not have reached genome-wide significance for other traits (and therefore be present in curated catalogues).

We use the term *summary-set* to refer to the minimum data required to perform 2SMR analysis - four columns comprising the SNP-exposure and SNP-outcome effects and standard errors, and rows corresponding to each SNP that is used as an instrument for the exposure.

At its most simple implementation, the regression of the SNP-exposure effect sizes against the SNP-outcome effect sizes, with greater weight afforded to those SNPs with smaller SNP-outcome standard errors, provides the estimate of the causal effect of the exposure on the outcome. This is known as the inverse-variance weighted (IVW) method.

@@

Here we adapt this approach to automatically filter SNPs that are liable to be invalid (Figure 1a). Other methods have been developed to identify invalid instruments for the purposes of exclusion from MR analyses [@Corbin2016; @Verbanck2017; @Zhu2017], based on the notion that outlying instruments are likely to be exhibiting horizontal pleiotropy. A potential drawback of this approach is that the outlier SNPs in a summary-set might be the only reliable ones. This could arise, for example, for the *CRP* variant instrumenting C-reactive protein levels [@Prins2016], or the *SLC2A9* variant influencing serum urate levels [@white2016plasma].

To avoid this, we primarily developed Steiger filtering to identify those instruments that are likely to be arising due to reverse cause, but it also has the potential to detect SNPs that are exhibiting horizontal pleiotropy or primarily influencing confounders. Further details are provided in the Online Methods.

@@

We consider the 14 MR methods described above, for which instruments can be supplied using two instrument selection strategies, leading to 28 methods in total (Supplementary table 2). In the context of this analysis, each method is considered to be an 'expert', taking a set of SNP-exposure and SNP-outcome effect sizes and their standard errors as inputs. Our objective is to select the expert most likely to be correct for a specific MR analysis. An overview of the approach is shown in Figure 2.

The mixture of experts method is a machine learning approach [@Jordan1994] which seeks to divide a parameter space into subdomains, such that a particular expert is used primarily for problems that reside in a subdomain most suited to that expert. In this case our objective is to identify characteristics of the SNP-exposure and SNP-outcome associations for which one specific MR method is most likely to yield highest statistical power for non-null associations, and lowest false discovery rates for null associations. This involves creating a 'gating function', whose purpose is to decide which expert to use for a specific MR analysis, given the parameter space that is occupied by that summary-set. The metrics are a mixture of regression diagnostics and are described in Supplementary table 3.

The gating function needs to be trained using data for which the true causal effect is known, and to this end we generated a large number of simulated summary-sets. We trained the gating function using random forest learning algorithms that seek to identify sectors of the parameter space that are most likely to return accurate estimates for a particular expert. Figure 2a illustrates how the gating function is trained using simulated data. New summary-sets can then be applied to the trained mixture of experts (MoE) model (Figure 2b). We call this implementation MR-MoE 1.0. Full details are provided in the Online Methods. MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR) [@Hemani2016].

The set of MR estimates obtained from this analysis, and the summary-sets used to generate them, are deposited in a Neo4j graph database. Here nodes representations exist for traits and SNPs; and relationship representations exist for SNP-trait associations and trait-trait MR estimates. Because each trait-trait association has up to 28 different MR estimates, for simplicity we also distill this down to a third relationship type comprising a single 'best estimate' for each trait-trait association using the following rules:

As statistical power for GWAS studies improves the likelihood of a significant association being discovered that doesn't act primarily on the trait of interest increases (Figure 1a). We evaluated the efficacy of the Steiger filtering approach for improving instrument selection using 100,000 simulated summary-sets comprising both null and non-null causal models. For each summary-set, instruments were selected based on the tophits strategy and the Steiger filtering strategy, and MR was performed using 14 different methods based on instruments selected from each of these strategies.

Following evidence that the Steiger filtering approach can improve on existing methods, we next hypothesised that a mixture of experts (MoE) model would be able to predict the most appropriate of the 28 MR methods and instrument selection strategies to apply to a particular summary-set based on its characteristics (Figure 2).

The ability to predict the performance for each of these methods is shown in Supplementary table 2. The prediction $R^2$ of whether a summary-set's status was truly null ($\beta_{xy}=0$) or non-null ($\beta_{xy} \neq 0$) against the method's prediction of the summary-set's status, ranged between 0.04 and 0.24. The method performance prediction was most effective for the Egger random effects model with Steiger filtering. The summary-set characteristics with the most importance for each of the predictors differed substantially between each summary-set, as well as the frequency for which each of the methods was selected in the testing summary-sets. The FDR of each method when chosen, compared to their averages across all summary-sets, reduced; and likewise the true positive rates of each method when chosen increased compared to their simulation-wide averages.

We compared the MoE performance in the simulations against each of the 28 methods, testing to see if it outperformed all other single methods. Figures 3a and 3b show that the MoE approach had the best general performance. Estimating the area under the receiver operator curve gave 0.84 for the MoE approach in terms of classifying the simulations as being null or non-null. Notably, the next best methods were median and mode based estimators using Steiger filtering. Under the assumption of widespread and diverse pleiotropic effects it is clear that all methods suffer from high false discovery rates (Figure 3a).

The trend of increasing GWAS sample sizes continues, but while the opportunity that this affords MR to be furnished with more instruments is typically welcomed, here we have shown that it is invalid instruments, exhibiting horizontal pleiotropy or reverse causation, that are more likely to be identified than valid ones. Strategising how MR is to be used in practice, therefore, must consider horizontal pleiotropy to be the rule rather than the exception. Following a recent flurry in development of MR methodology, we have devised a machine learning approach, MR-MoE, that seeks to predict the performance of each MR method, selecting the one which is most likely to maximise power whilst minimising FDR for a specific summary-set. The mixture of experts approach that we present here is trained using random forest decision trees applied to extensive and diverse simulations, and we demonstrate that it makes substantial performance improvements over any other single method in the presence or absence of extensive horizontal pleiotropy. As such, our method contributes to the field of meta-learning but also to the rapidly growing field of automatic model selection and hyper-parameter optimization in machine learning [@kotthoff2016auto].

One advantage to generating a method that consolidates across many methods is that it can prevent cherry picking the most appealing method based on their results. But in addition to correctly handling horizontal pleiotropy, there are many other limitations that prevent MR from being a panacea for causal inference. Many of the associations are biologically impossible, for example where early life phenotypes appear to be influenced by later stage phenotypes. Though these associations can be informative, their interpretations as causal relationships are far from clear. Similarly, often disease traits appear to causally relate to other phenotypes, but GWAS is typically performed on the liability scale, hence the causal estimate reflects not the presence or absence of disease, but the underlying risk of disease. Again, interpretation of such associations can be problematic.

We also encounter a potential new problem in using machine learning approaches to infer the correct model for a particular summary-set, namely, that the data is choosing the model. This is liable to contribute to elevated type 1 error rates, even though we have attempted to separate the information used for optimisation from the information used to predict performance. The extent to which a machine based method selector suffers from this more than a human manually inferring the result most likely to be true is not clear. Though MR-MoE does exhibit higher type 1 error rates than are desirable, they still remain amongst the lowest compared to all other MR methods that do not suffer from this issue.

## Acknowledgements

This work was supported by the UK Medical Research Council (MC_UU_12013/1).

## Online methods

### Steiger filtering

An approach to inferring the causal direction between phenotypes recently developed [@Hemani2017] uses the following basic premise. If trait $A$ causes trait $B$ then $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$ because the $cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}$. This will be true under most circumstances, but some parameters of measurement error or unmeasured confounding can lead to the inequality reversing direction, as has been explored previously [@Hemani2017].

The Steiger test is applied to each variant in turn and we exclude any $g_{A}$ for which $cor(g_{i}, A)^2 > cor(g_{i}, B)^2$, indicating that it is unlikely to primarily associate with $B$ relative to $A$. Similarly, for SNPs that influence confounders of $A$ and $B$ or exhibit horizontal pleiotropy, the difference in $cor(g_{i}, A)^2$ and $cor(g_{i}, B)^2$ will be reduced, increasing the likelihood of the SNP being excluded because the Steiger Z-test is less likely to be significant. Hence we also exclude any $g_{A}$ for which the $cor(g, x)$ and $cor(g, y)$ are not significantly different at an arbitrary threshold of $p > 0.05$.

To estimate $cor(g, x)^2$, if $x$ is continuous we obtain the F-statistic from the reported p-value and sample size and then $cor(g, x)^2 = \frac{F}{N - 2 - F}$. If $x$ is binary then we estimate the variance of the underlying liability explained by the SNP, $cor(g, x)^2 = \frac{V_a}{V_a + V_e}$. Here, $V_e = \pi^2/3$, and $V_a = 2\beta^2f(1-f)$, where $\beta$ is the log odds ratio and $f$ is the allele frequency of the SNP in the population [@Lee2013c]. $f$ can be estimated using the allele frequency of the SNP in an ascertained sample by deriving the $2 \times 2$ contingency table from the odds ratio $e^\beta$, allele frequency in the ascertained sample $f_{cc}$, and number of cases $N_1$ and controls $N_0$. A Z-score to test the hypothesis that $cor(g, x)$ and $cor(g, y)$ are significantly different can be generated using the Steiger test for correlated correlations.

### Mixture of experts implementation for MR

The mixture of experts approach seeks to use simulations to learn the circumstances in which each of the 28 MR methods is likely to perform most accurately. Having trained the model on simulations, new summary-sets can be applied to obtain causal estimates from the method that is predicted to be the most reliable for that summary-set.

It should be noted that in the original hierarchical mixture of experts approach of [@Jordan1994] the gating function and the experts share the same input space and are trained simultaneously using expectation-maximisation. In our approach the gating function is defined over a separate input space consisting of the parameter space of the experts, and is trained separately in a supervised setting using simulated data. As such our MR-MoE approach fits the framework of meta-learning [@brazdil2008metalearning; @smith2009cross; @lemke2015metalearning] where machine learning methods are applied at the 'meta-level' to analyse the results of machine learning experiments at the 'base-level', in order to be able to understand and model the capabilities of each expert and recommend which expert should be applied to a given problem. From this perspective the 28 MR methods are the base-models, the gating function is the meta-model, and its meta-features are the parameters of the summary-sets.

@@

The MoE is trained using summary-sets generated from simulations (Figure 2a). Each *summary-set* can be fed into any of the 28 experts to obtain MR causal effects. The simulations used to generate these summary-sets seek to cover a range of pleiotropic scenarios, including where some proportion of SNPs exhibit directional or balanced horizontal pleiotropy, or where SNPs influence confounding variables.

We simulate two individual level datasets for which there are $N_x$ and $N_y$ samples, and $M$ SNPs, where each SNP has effect allele frequency of $f_m \sim U(0.05, 0.95)$. These datasets are used to obtain the SNP effects for the exposure trait $x$ and the outcome trait $y$, respectively, to create summary-sets, using the following sampling criteria:

@@

For each simulation we used linear regression to estimate the genetic effect of each SNP $M$ on $x$ in sample 1, and each SNP $M$ on $y$ in sample 2. We then perform MR analysis in both directions, mimicking GWAS by retaining SNPs that have $p < 5e-8$ in sample 1 to perform MR of $x$ on $y$ (the true causal direction for non-null simulations), and retaining SNPs that have $p < 5e-8$ in sample 2 to perform MR of $y$ on $x$ (the reverse causal direction for non-null simulations). We treat the summary data (effect sizes and standard errors) used for estimating $x \rightarrow y$ the summary data used for estimating $y \rightarrow x$ as two separate summary-sets. Hence, for each simulation two summary-sets are generated which are analysed to produce 28 MR estimates each. We performed 100,000 simulations using these parameters, resulting in 200,000 summary-sets.

@@

We aim to maximise statistical power for summary-sets where $\beta_{x,y} \neq 0$ and minimise false discovery rates for summary-sets where $\beta_{x,y} = 0$. To train random forest decision trees to predict performance for a particular method $h(O_{w,d}, \textbf{z}_{d})$ is generated where the training set of input metrics for summary-set $d$ is $\textbf{z}_{d}$ and the response (optimisation function) is

@@

where $p_{m,d}$ is the p-value for method $m$ on summary-set $d$.

@@

For each training summary-set we record a set of 53 metrics $\textbf{z}_{d}$ (Supplementary table 3), and an outcome $O_{w,d}$, which is a measure of how well that method performed for each particular simulated summary-set. For each of our 28 methods, we need to create a model that predicts the performance of the method based on metrics generated from a summary-set. To do this, for each method we train a random decision forest to predict that method's performance using the summary-set's metrics. The random forest approach is well suited to this problem because there are likely to be non-continuous combinations of different metrics that improve on prediction over, for example, a simple linear model that does not learn about interactions.

As a simple hypothetical example - if a summary-set exhibits a single outlier but is otherwise exhibiting no heterogeneity then the following methods could arguably perform well:

@@

Having generated random forest decision trees for each of the 28 methods using 133,000 of the simulations, we then applied them to the remaining 67,000 summary-sets to predict which method would have the highest performance for each of the remaining summary-sets. Finally we compare the performance of the method selected by the MoE against all remaining methods. The default settings for the `randomForest` package in R [@Liaw2002] were used to train the models. MR-MoE 1.0 is implemented in the TwoSampleMR R package available at [github.com/MRCIEU/TwoSampleMR](https://github.com/MRCIEU/TwoSampleMR) [@Hemani2016].
